---
title: "FotW collections on GBIF — Summary"
author: "Sven Buerki - Boise State University"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE, cache.path = "cache_summary/")
library(dplyr)
library(knitr)
library(kableExtra)
```

```{r load_data}
# GBIF occurrences
occurrence_data <- read.table(
  "GBIF_data/0000210-250515123054153/occurrence.txt",
  sep = "\t", header = TRUE, quote = "", comment.char = "", stringsAsFactors = FALSE)
occurrence_data$eventDate        <- as.Date(occurrence_data$eventDate, format = "%Y-%m-%d")
occurrence_data$decimalLatitude  <- as.numeric(occurrence_data$decimalLatitude)
occurrence_data$decimalLongitude <- as.numeric(occurrence_data$decimalLongitude)
occurrence_data[occurrence_data == ""] <- NA

occurrence_data_plants <- subset(occurrence_data, kingdom == "Plantae")

# Dataset metadata
dataset_info <- read.csv("GBIF_data/datasets_download_usage_0000210-250515123054153.tsv", sep = "\t")

# Multimedia
multimedia_data <- read.table(
  "GBIF_data/0000210-250515123054153/multimedia.txt",
  sep = "\t", header = TRUE, quote = "", comment.char = "", stringsAsFactors = FALSE)
images_plants <- subset(multimedia_data, gbifID %in% occurrence_data_plants$gbifID)
images_plants$rightsHolder[images_plants$rightsHolder == ""] <- "No_rightsHolder"
images_plants$license[images_plants$license == ""]           <- "No_license"

# FotW occurrences
FotW_data <- read.csv("FotW_DB/occurrences.csv", stringsAsFactors = FALSE)
FotW_data$evenDateFormatted    <- sapply(strsplit(FotW_data$eventDate, " "), `[[`, 1)
FotW_data$recordNumberFormatted <- gsub("[^0-9]", "", FotW_data$recordNumber)
```

---

## Q1 — How many occurrences are on GBIF?

After filtering to kingdom **Plantae**, Christopher Davidson's GBIF download contains
**`r nrow(occurrence_data_plants)` occurrences**.

Each occurrence is directly accessible at `https://www.gbif.org/occurrence/<gbifID>`.

---

## Q2 — Which institutions supply data to GBIF?

```{r q2}
institutions <- as.data.frame(table(occurrence_data_plants$datasetKey))
colnames(institutions) <- c("DatasetKey", "N_occurrences")
institutions <- institutions[order(as.numeric(institutions$N_occurrences), decreasing = TRUE), ]
institutions$Dataset_Title <- dataset_info$dataset_title[
  match(institutions$DatasetKey, dataset_info$dataset_key)]

kable(institutions[, c("Dataset_Title", "N_occurrences")],
      col.names = c("Dataset / Institution", "N occurrences"),
      row.names = FALSE)
```

> The vast majority of occurrences come from **TROPICOS** and **RSA - California Botanic Garden Herbarium**
> (`r sum(institutions$N_occurrences[1:3])` combined).
> TROPICOS also acts as a data aggregator: `r institutions$N_occurrences[which(institutions$Dataset_Title == "Tropicos Specimens Non-MO")]` occurrences are listed there but not physically deposited at MO.

---

## Q3 — Where are physical specimens deposited?

```{r q3}
occurrence_data_plants$institutionCode[is.na(occurrence_data_plants$institutionCode)] <- "No_institution"

top20 <- sort(table(occurrence_data_plants$institutionCode), decreasing = TRUE)[1:20]
top20_df <- as.data.frame(top20)
colnames(top20_df) <- c("Institution", "N_occurrences")

kable(top20_df, row.names = FALSE,
      caption = "Top 20 institutions holding physical specimens.")
```

> **MO, RSA, NY and SRP** are the main institutions housing FotW specimens.

---

## Q4 — Are digital images available online?

Of `r nrow(occurrence_data_plants)` GBIF plant occurrences,
**`r nrow(images_plants)` (`r paste0(round(100 * nrow(images_plants) / nrow(occurrence_data_plants), 1), "%")`) have digital images** linked on GBIF.

```{r q4_license}
lic <- as.data.frame(sort(table(images_plants$license), decreasing = TRUE))
colnames(lic) <- c("License", "Count")
kable(lic, row.names = FALSE, caption = "Licenses for images.")
```

```{r q4_rights}
rh <- as.data.frame(sort(table(images_plants$rightsHolder), decreasing = TRUE))
colnames(rh) <- c("Rights holder", "Count")
kable(head(rh, 15), row.names = FALSE, caption = "Top 15 rights holders for images.")
```

> Note: **SRP (Boise State)** does not currently provide image URLs to GBIF;
> images for SRP specimens are available via the
> [PNW Herbaria portal](https://www.pnwherbaria.org/) (see Q8).

---

## Q5 — How many unique collections are represented?

```{r q5}
collection_id <- paste(occurrence_data_plants$recordedBy,
                       occurrence_data_plants$recordNumber,
                       occurrence_data_plants$eventDate, sep = "_")
unique_col <- as.data.frame(table(collection_id))
```

| Metric | Value |
|---|---|
| Unique collections | `r nrow(unique_col)` |
| Occurrences missing record number | `r sum(is.na(occurrence_data_plants$recordNumber))` |

> Multiple GBIF occurrences per collection indicate **duplicate specimens** distributed across herbaria.

---

## Q6 — How many taxa and from where?

```{r q6}
n_taxa    <- length(unique(occurrence_data_plants$scientificName))
n_families <- length(unique(occurrence_data_plants$family[!is.na(occurrence_data_plants$family)]))
n_genera  <- length(unique(occurrence_data_plants$genus[!is.na(occurrence_data_plants$genus)]))
n_species <- length(unique(occurrence_data_plants$species[!is.na(occurrence_data_plants$species)]))

continent_ct <- as.data.frame(sort(table(occurrence_data_plants$continent), decreasing = TRUE))
colnames(continent_ct) <- c("Continent", "N_occurrences")

country_ct <- as.data.frame(sort(table(occurrence_data_plants$countryCode), decreasing = TRUE))
colnames(country_ct) <- c("Country code", "N_occurrences")
```

**Taxonomy:**

| Level | Count |
|---|---|
| Scientific names (all ranks) | `r n_taxa` |
| Families | `r n_families` |
| Genera | `r n_genera` |
| Species | `r n_species` |

**Geography:**

```{r q6_continent}
kable(continent_ct, row.names = FALSE, caption = "Occurrences by continent.")
```

```{r q6_country}
kable(head(country_ct, 20), row.names = FALSE,
      caption = "Top 20 countries by occurrence count.")
```

---

## Q7 — Other occurrence information (phenology, conservation)

```{r q7}
pheno <- as.data.frame(sort(table(occurrence_data_plants$reproductiveCondition), decreasing = TRUE))
colnames(pheno) <- c("Reproductive condition", "Count")
pheno <- pheno[pheno$`Reproductive condition` != "", ]

iucn <- as.data.frame(sort(table(occurrence_data_plants$iucnRedListCategory), decreasing = TRUE))
colnames(iucn) <- c("IUCN category", "Count")
iucn <- iucn[iucn$`IUCN category` != "", ]
```

**Phenology** (`reproductiveCondition` field):

```{r q7_pheno}
if (nrow(pheno) > 0) {
  kable(pheno, row.names = FALSE)
} else {
  cat("No reproductive condition data available in this dataset.")
}
```

**IUCN Red List status** (`iucnRedListCategory` field):

```{r q7_iucn}
if (nrow(iucn) > 0) {
  kable(iucn, row.names = FALSE)
} else {
  cat("No IUCN Red List data available in this dataset.")
}
```

---

## Q8 — Linking FotW images with SRP specimens

SRP specimen images are not on GBIF but are accessible via the
[Consortium of Pacific Northwest Herbaria](https://www.pnwherbaria.org/).
The local file `Consortium_PacificNorthWest_herbaria/CPNWH_20250520-112502.csv`
lists all Davidson SRP collections with their accession numbers and an `Imaged?` flag.

Image URL pattern (accession must be zero-padded to 6 digits):

```
https://www.pnwherbaria.org/images/jpeg.php?Image=SRP<ACCESSION>.jpg
# Example: https://www.pnwherbaria.org/images/jpeg.php?Image=SRP034726.jpg
```

---

## Q9 — Overlap between FotW and GBIF

Records are matched on a composite key: **numeric record number + collection date**
(e.g., `12687_2014-10-04`).

```{r q9_match}
rn_fotw   <- paste(FotW_data$recordNumberFormatted, FotW_data$evenDateFormatted, sep = "_")
rn_gbif   <- paste(occurrence_data_plants$recordNumber, occurrence_data_plants$eventDate, sep = "_")

gbif_matched <- occurrence_data_plants[rn_gbif %in% rn_fotw, ]
gbif_matched$match_key <- rn_gbif[rn_gbif %in% rn_fotw]

FotW_data$match_key <- rn_fotw
fotw_dedup <- FotW_data[!duplicated(FotW_data$match_key),
                        c("match_key", "occurrenceID")]
colnames(fotw_dedup)[2] <- "FotW_occurrenceID"

gbif_matched <- merge(gbif_matched, fotw_dedup, by = "match_key", all.x = TRUE)
gbif_matched$gbifURL <- paste0("https://www.gbif.org/occurrence/", gbif_matched$gbifID)
```

**`r nrow(gbif_matched)` GBIF occurrences** correspond to **`r length(unique(gbif_matched$FotW_occurrenceID))` unique FotW records**.

```{r q9_inst}
inst_match <- as.data.frame(sort(table(gbif_matched$institutionCode), decreasing = TRUE))
colnames(inst_match) <- c("Institution", "N_occurrences")
kable(inst_match, row.names = FALSE,
      caption = "Institutions holding specimens present in both GBIF and FotW.")
```

```{r q9_country}
cc_to_name <- c(EC="Ecuador", US="United States", NC="New Caledonia", FJ="Fiji",
                PE="Peru", CH="Switzerland", MG="Madagascar", BO="Bolivia",
                PY="Paraguay", GA="Gabon", BZ="Belize", ZA="South Africa",
                VN="Vietnam", CN="China", SI="Slovenia")

country_match <- as.data.frame(sort(table(gbif_matched$countryCode), decreasing = TRUE))
colnames(country_match) <- c("Country_code", "N_occurrences")
country_match$Country <- cc_to_name[as.character(country_match$Country_code)]
country_match <- country_match[, c("Country", "Country_code", "N_occurrences")]

kable(country_match, row.names = FALSE,
      caption = "Countries of collection for occurrences present in both GBIF and FotW.")
```

### Full list of matched collections

```{r q9_table}
display <- gbif_matched[, c("gbifURL", "FotW_occurrenceID", "scientificName",
                             "family", "recordNumber", "eventDate", "recordedBy",
                             "institutionCode", "catalogNumber",
                             "countryCode", "stateProvince", "locality")]
display <- display[order(display$family, display$scientificName), ]
colnames(display) <- c("GBIF URL", "FotW ID", "Scientific name", "Family",
                        "Record no.", "Date", "Recorded by",
                        "Institution", "Catalog no.",
                        "Country", "State/Province", "Locality")
kable(display, row.names = FALSE)
```

```{r q9_export, include=FALSE}
write.csv(gbif_matched, "GBIF_FotW_matched_collections.csv", row.names = FALSE)
```
